---
title: "R Workshop: Analyzing MPR survey responses using 'final' poll format" 
author: "Amelia Ritger"
date: "11/07/2022"
output: html_document
---

## Before starting, you need to convert .xlsx responses to .csv! Bath convert following these directions:
Download LibreOffice, open up data folder in terminal, then run this script:
/Applications/LibreOffice.app/Contents/MacOS/soffice --headless --convert-to csv *.xlsx

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

install.packages("tidyverse", "janitor", "here", "scales", "viridis", "ggimage") #you only need to do this once

library(tidyverse) #use for so many things!
library(janitor) #use to clean up column headers
library(here) #use to work with folder structure
library(scales) #use to customize figure axes
library(viridis) #use to apply colorblind friendly palette
library(ggimage) #use to plot human icon
```

## Import Data
```{r, message=FALSE}
#Create a list of all .csv files we have received
file_names <- list.files(here("survey-analysis", "ocnms-responses"), pattern = ".csv$", recursive = TRUE, full.names = TRUE)

#Take that list and merge all .csv files into one dataframe
all_merged <- read_csv(file_names, col_names = FALSE) %>%
  select(1:5) %>% #only keep first 5 columns
  setNames(c("affiliation", "focus_area", "description", "issues_to_rate", "score")) #change column names, use snake case (or use clean_names)

#Remove header rows and OTHER rows
all_clean <- all_merged %>%
  filter(is.na(score) | !str_detect(score, 'Focus Area'), #remove Focus Area header
         is.na(score) | !str_detect(score, 'Priority'), #remove Priority Topic header
         is.na(score) | !str_detect(focus_area, "[:digit:]"), #remove header rows (those that contain (number))
         is.na(score) | !str_detect(score, 'Score'), #keep non-ranked issues, just in case - especially for comments and "other" responses
         !is.na(description)) %>% #remove blank rows in Description
  mutate(issues_to_rate = ifelse(str_detect(issues_to_rate, 'Water quality') == TRUE, "Water quality protection", issues_to_rate)) #something is weird with the value of "water quality protection", so fix that

rankings <- all_clean %>%
  mutate(score = replace_na(score, "None")) %>% #if someone left an issue blank in the rating, rename it "None" so we can assign a score of 0
  filter(str_detect(score, 'High|Medium|Low|None')) %>% #remove "do not score" rows (aka OTHER, comments)
  mutate(value = ifelse(score == "Low", 1, ifelse(score == "Medium", 2, ifelse(score == "High", 3, 0)))) %>% #convert scores to numeric: high = 3, medium = 2, low = 1, no response/NA = 0
  filter(!is.na(issues_to_rate),
#remove rows with "NA" in issues to rate column
         !str_detect(issues_to_rate, 'Write')) #remove "Other" rows

#filter out the comments
comments <- all_clean %>%
  filter(str_detect(description, 'Other?'), #keep "Other" rows
         !str_detect(score, 'High|Medium|Low') | is.na(score)) #people weren't supposed to score "other", so remove rows with a score
```

## Group and count total responses
```{r}
issue_totals <- rankings %>%
  group_by(issues_to_rate) %>% #group by issue
  summarize(total = sum(value), #total score for each issue
            mean = mean(value), #mean
            sd = sd(value), #standard deviation
            n = n()) %>% #number of responses
  arrange(desc(total)) #arrange from highest score > lowest score 
```

## Plot mean and sd of responses, including total score
```{r}
ggplot(data = issue_totals) +
  geom_col(aes(x=reorder(issues_to_rate, mean), y=mean, fill = mean)) +
  geom_errorbar(aes(x = issues_to_rate, ymin=mean, ymax=mean+sd), colour="black", width=0.5) +
  geom_text(aes(x=reorder(issues_to_rate, -mean), y=-0.15, label = total)) + #add total counts
  scale_fill_viridis(direction = -1) + #add color for fun
  labs(x="", y="Average Ranking", caption = "*Bar sizes and color represent average (mean) ranking. Error bars represent standard deviation. Numbers represent total (sum) score.") + #rename axis titles
  coord_flip() +
  theme_minimal() +
  theme(legend.position = "none",
        plot.caption = element_text(hjust = 5, face = "italic"))

#ggsave(here("figures", "mean_sd.png"), height=15, width=25, units="cm", bg = "white")
```

Plot up the level of consensus AKA s.d.
```{r}
issue_totals <- issue_totals %>%
  mutate(sd2 = (1-sd),
         issues_to_rate = fct_reorder(issues_to_rate, desc(sd)))

ggplot(issue_totals, aes(x=issues_to_rate, y=sd2)) +
  geom_segment(aes(x=issues_to_rate, xend=issues_to_rate, y=0, yend=sd2, color = mean)) +
  geom_image(aes(image="https://upload.wikimedia.org/wikipedia/commons/d/d8/Person_icon_BLACK-01.svg", color = mean), size=0.03) +
  scale_color_viridis(direction = -1, name = "Mean rank") +
  scale_y_continuous(breaks=c(0,0.5,1), labels=c("Low", "Medium", "High")) +
  labs(x = "", y = "Group consensus on ranking", caption = "*Consensus was assessed from standard deviation for each issue. Color represents average ranking for each issue.") +
  coord_flip() +
  theme_minimal() +
  theme(
    axis.ticks.y = element_blank(),
    plot.caption = element_text(hjust = 5, face = "italic"))

#ggsave(here("figures", "consensus.png"), height=15, width=25, units="cm", bg = 'white')
```

## Break down each issue by ranking from each seat, with heatmap
```{r, message = FALSE}
rankings_count <- rankings %>%
  group_by(issues_to_rate) %>%
  mutate(count = row_number()) %>% #get the number of consecutive counts for each issue (to plot)
  ungroup() %>%
  mutate(issues_to_rate = fct_inorder(issues_to_rate), #set order of issues as order in which they appear on data table
         score = fct_relevel(score, c("High", "Medium", "Low"))) #set order of scores (not alphabetical)

ggplot(rankings_count, aes(x=reorder(issues_to_rate, desc(issues_to_rate)), y=count, fill=score)) + #use count and not affiliation because everyone at OCNMS has the same affiliation
  geom_tile(color = "black") +
  coord_flip() +
  labs(x="", y="") +
  scale_y_discrete(position = "right") +
  scale_fill_viridis(discrete = TRUE) +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=45, vjust = -1, hjust=0))

#ggsave(here("figures", "individual_responses_heatmap.png"), height=15, width=25, units="cm", bg = "white")
```

Save the comments as a .csv
```{r}
comments <- comments %>%
  arrange(focus_area) %>%
  select(1:4)

#write_csv(comments, here("figures", "comments.csv"))
```

## Another way to visualize "Convergence via selecting example topics
```{r}
rankings_consensus <- rankings_count %>%
  filter(issues_to_rate %in% c("Community outreach", "Canadian government", "Changing ocean conditions")) %>% #keep only topics of interest (good examples of range of rankings/consensus)
  mutate(issues_to_rate = fct_relevel(issues_to_rate, c("Community outreach", "Canadian government", "Changing ocean conditions"))) #set order for figure

ggplot(data = rankings_consensus, aes(x=issues_to_rate, y=count)) + 
  geom_image(aes(image="https://upload.wikimedia.org/wikipedia/commons/d/d8/Person_icon_BLACK-01.svg", color = score), size=0.1) + #use human icon
  geom_text(label = "'High' average rank, 'High' convergence (Mean =  3.00, Standard deviation = 0.00)", x=3.35, y=2, size = 3.5) + #label for Changing ocean conditions
  geom_text(label = "'Low' average rank, 'Medium' convergence (Mean = 1.21, Standard deviation = 0.43)", x=2.35, y=2.2, size = 3.5) + #label for Canadian government
  geom_text(label = "'Medium' average rank, 'Low' convergence (Mean =  2.29, Standard deviation = 0.92)", x=1.35, y=2.2, size = 3.5) + #label for Community outreach
  scale_color_viridis(discrete=TRUE, name = "Rank") + #make each icon the rank from each respondent
  coord_flip(clip="off") + #flip coordinates to make it more readable, clip = 'off' so you can have geom_text anywhere
  labs(y="", x="", caption = "*Each icon represents a OCNMS staff person's response. Color represents the staff person's ranking of the topic. Convergence was calculated from the standard deviation of rankings.") + #add descriptive caption, remove x and y axis labels
  theme_classic() + #simplify the theme
  theme(axis.ticks.x = element_blank(), #remove x axis ticks
        axis.text.x = element_blank(), #remove x axis text
        axis.line = element_blank(), #remove axes lines
        legend.position = "bottom", #put legend at the bottom
        plot.caption = element_text(hjust = 0, face = "italic"))

#ggsave(here("figures", "sac-convergence-example.png"), height=15, width=25, units="cm", bg = "white")
```
